Gelation as a condensation frustrated by hydrodynamics and mechanical tension

### 2012-06-20 paper.tex
Start discussing spinodal Fleer2008

We use a novel experimental setup that allows the observation of the entire process of gelation at a single particle level (see Methods). The principle is to introduce salts through a membrane in a charged colloids and non adsorbing polymers suspension (see Fig.~\ref{fig:methods}a). The polymers induce a very short range depletion attraction between colloids, however long range electrostatic repulsion between the colloids prevent any aggregation. When we introduce salt the charges get screened and attraction suddenly kicks in, eventually leading to the formation of a gel. The whole process can take place on the microscope stage and can thus be observed in three dimensions from the very beginning.

Figure~\ref{fig:methods}c-h shows a computer reconstruction from experimental coordinates of a typical gelation experiment. We are able to follow the state of the system from a Wigner crystal (Fig.~\ref{fig:methods}c) where the particles are far apart due to repulsion to a fully percolated network (Fig.~\ref{fig:methods}h).

\section{Cluster aggregation} % Subject I: Cluster formation
% Topic sentence
Figure~\ref{fig:methods}a shows the phase diagram of all samples.
First of all, we focus on cluster formation since clusters constitute " building blocks " of gels.
% Results of birth
Figure~\ref{fig:fluid_dilute} a shows the compaction dynamics of the clusters with three colloids in the early stage.
The results show that the clusters can hardly accomplish compaction within one Brownian motion time and clusters with four and five colloids show very similar tendency (Supplement).
% Explanation 1) Local energy minimization.
If one assumes that each colloid can move independently by Brownian motion and static short-range attractions are induced between colloids, the clusters with three colloids should reach the compact states within one Brownian motion time.
This explanation contradicts to our results.
% Addressing our question
In other words, some interactions stabilize linear structures, competing with the principle of energy minimization.
% Explanation 2) Electrostatic repulsion.
Although the effect of electrostatic repulsions has been studied, our results are observed after salt screens surface charges and electrostatic repulsions can be negligible in our system.
% Explanation 3) Hydrodynamic interactions.
Thus, the only possible explanation is the effect of hydrodynamic interactions.
Hydrodynamic interactions originate from the incompressivility of the solvent, whose effect between colloids have been demonstrated by both numerical simulations and the experiments of optical tweezers.
For compaction, clusters must squeeze the solvent inside the cluster to the outside.
% Conclusion.
Our results suggest that squeezing process can be the bottleneck of compaction and hydrodynamic interactions enhance linear structures in clusters.

@@

From direct confocal measurement~\cite{Royall2007, Poon2012}, we estimate the hard-core diameter of our colloids ($\sigma=\unit{2.75}{\micro\metre}$) and the range of the interaction potential (that confirmed our scaling of $R_g$ within $1\%$), leading to a polymer-colloid size ratio $q = 2R_g/\sigma = 0.10(6)$. Two particles are considered bonded when their centres are within $\sigma+2R_g$, which corresponds to the first minimum of $g(r)$.

>>>>

We use a novel experimental set-up that allows the observation of the entire process of gelation at a single particle level (see Methods). In a suspension of sticky colloids with long range repulsion, we suddenly turn off the repulsion. The short range attraction become dominant, eventually leading to the formation of a gel. The whole process takes place on the microscope stage and can thus be observed in three dimensions from the very beginning.

Figure~\ref{fig:methods}c-h shows a computer reconstruction from experimental coordinates of a typical gelation experiment. We are able to follow the state of the system from (Fig.~\ref{fig:methods}c) a Wigner crystal where the particles are far apart due to repulsion to (Fig.~\ref{fig:methods}h) a fully percolated network. Using this protocol, we explored the phase diagram (Fig.~\ref{fig:methods}b) and found three possible behaviours: permanent gelation as above, transient gelation and cluster formation. Transient gelation is characterised by the formation of a percolating network that melts into separate clusters over a few tens of Brownian times (the time needed for an isolated particle to diffuse over its own radius, here about $\unit{10}{second}$).

We observe in Fig.~\ref{fig:methods}b that the spinodal line predicted by the generalized free volume theory~\cite{Fleer2008} corresponds roughly to the boundary between permanent and transient gelation. Furthermore, it means that in our system gelation can happen transiently without an underlying spinodal decomposition dynamics, somewhat in contradiction to~\cite{lu2008gelation}.

Our protocol is not equivalent to a redispersion of an existing gel at the same state point either by shaking or shear melting like in  e.g.~\cite{lu2008gelation,Teece2011,Bartlett2012}. It is closer to a quench and to application relevant processes where the sudden addition of a component (e.g. salt, acid, depletion agent) induces aggregation in an otherwise stable suspension (e.g. milk).

\section{Cluster aggregation} % Subject I: Cluster formation
% Topic sentence
Figure~\ref{fig:methods}a shows the phase diagram of all samples.
First of all, we focus on cluster formation since clusters constitute ``building blocks'' of gels.
% Results of birth
Figure~\ref{fig:fluid_dilute} a shows the compaction dynamics of the clusters with three colloids in the early stage.
The results show that the clusters can hardly accomplish compaction within one Brownian motion time and clusters with four and five colloids show very similar tendency (Supplement).
% Explanation 1) Local energy minimization.
If one assumes that each colloid can move independently by Brownian motion and static short-range attractions are induced between colloids, the clusters with three colloids should reach the compact states within one Brownian motion time.
This explanation contradicts to our results.
% Addressing our question
In other words, some interactions stabilize linear structures, competing with the principle of energy minimization.
% Explanation 2) Electrostatic repulsion.
Although the effect of electrostatic repulsions has been studied, our results are observed after salt screens surface charges and electrostatic repulsions can be negligible in our system.
% Explanation 3) Hydrodynamic interactions.
Thus, the only possible explanation is the effect of hydrodynamic interactions.
Hydrodynamic interactions originate from the incompressivility of the solvent, whose effect between colloids have been demonstrated by both numerical simulations and the experiments of optical tweezers.
For compaction, clusters must squeeze the solvent inside the cluster to the outside.
% Conclusion.
Our results suggest that squeezing process can be the bottleneck of compaction and hydrodynamic interactions enhance linear structures in clusters.

@@

From direct confocal measurement~\cite{Royall2007, Poon2012}, we estimate the hard-core diameter of our colloids ($\sigma=\unit{2.75}{\micro\metre}$) and the range of the interaction potential (that confirmed our scaling of $R_g$ within $1\%$), leading to a polymer-colloid size ratio $q = 2R_g/\sigma = 0.10(6)$. Spinodal and binodal lines on Fig.~\ref{fig:methods}b are calculated from this size ratio using the generalized free volume theory~\cite{Fleer2008}. Two particles are considered bonded when their centres are within $\sigma+2R_g$, which corresponds to the first minimum of $g(r)$.
